---
title: "ggplot2 Graphics Tutorial"
author: "Terra Dressler"
date: "November 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

In this tutorial, we will explore functions within ggplot2 that allow us to customize essentially every aspect of how our plots look.

We will use dataset of penguin measurements from Palmer Archipelago, Antarctica. These data are originally from this publication:
https://doi.org/10.1371/journal.pone.0090081 and the palmerpenguins package was created by Allison Horst at UCSB. For more information about this dataset and package, check out https://allisonhorst.github.io/palmerpenguins/

We will make:

  1. a scatter plot
  
  2. a boxplot
  
  3. a point range plot
  
We will manipulate aesthetics by:

  1. Using basic ggplot commands to change point size, shape, transparency, outline 
    
  2. Exploring several different options for changing plot colors
    
  3. Changing axis scales, labels, font sizes, and legend appearance
    
  4. Adding outputs of some basic statistical tests to our plot
    
  5. Customizing our plot theme

###1. Load Packages
```{r}
 
library(tidyverse) 
library(palmerpenguins) #contains the penguin data!
library(RColorBrewer) #contains color palettes we can can call on
library(ggsci) #another color package, one of my favorites!
library(scales) #will use to view color scales and hex codes for ggsci palettes
library(ggpubr) #will use to add some statistical outputs to our plots

```

Note: some other great color packages include viridis, dichromat, ggthemes, ghibli, wesanderson, pals, and rcartocolor

###2. Read in our "penguins" dataset

```{r}

penguins <- palmerpenguins::penguins %>% 
  rename(flipper = flipper_length_mm, bill_length = bill_length_mm, bill_depth = bill_depth_mm) %>% 
  mutate(mass_kg = body_mass_g/1000, bill_ratio = bill_length/bill_depth) %>% 
  na.omit() 
  
```



###3. Create a scatter plot 
####Say we are interested in how flipper length scales as a function of body mass for each penguin species. We will first make a basic scatter plot to visualize this relationship
```{r}

ggplot(penguins, aes(x = mass_kg, y = flipper))+
  geom_point()

```

####Default plot output is not great. We want to visualize each species separately with trendlines and make our plot easier to interpret and prettier to look at!

    
```{r}
#1. Basic commands, theme changes
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species, shape = species))+ 
  geom_point(alpha = 0.8, size = 2)+ #can change shape and stroke (shape outline size) here
  theme_classic() #show a couple other themes

#2. Axis labels
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species))+ 
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  labs(y = expression("Flipper Length (mm)", x = "Body Mass (kg)")) #mention title and caption too

#3. Degrees, subscripts, and superscripts within labels 
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species))+ 
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  labs(y = expression("Oxygen Uptake Rate  " (mg*O[2] * min^-1)), x = expression("Temperature " ( degree*C))) #example axis titles with subscripts, etc

#4. Axis scales and breaks
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species))+ 
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  labs(y = "Flipper Length (mm)", x = "Body Mass (kg)")+
  scale_y_continuous(limits = c(160, 240), breaks = c(160, 180, 200, 220, 240))+
  scale_x_continuous(limits = c(2.5,6.5))

#5. Colors!! Part 1: R colors
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species))+ 
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  labs(y = "Flipper Length (mm)", x = "Body Mass (kg)")+
  scale_y_continuous(limits = c(160, 240), breaks = c(160, 180, 200, 220, 240))+
  scale_x_continuous(limits = c(2.5,6.5))+
  scale_color_manual(values = c("darkorange", "darkorchid", "cyan4")) #See R colors cheatsheet for color names. Can also use color hex codes here! Match to powerpoint!

#6. Colors!! Part 2: RColorBrewer 
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species))+ 
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  labs(y = "Flipper Length (mm)", x = "Body Mass (kg)")+
  scale_y_continuous(limits = c(160, 240), breaks = c(160, 180, 200, 220, 240))+
  scale_x_continuous(limits = c(2.5,6.5))+
  scale_color_brewer(palette = "Dark2") #see R color cheatseet for palette names OR:

display.brewer.all()
display.brewer.all(colorblindFriendly = T)

#7. Colors! Part 3: ggsci
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species))+ 
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  labs(y = "Flipper Length (mm)", x = "Body Mass (kg)")+
  scale_y_continuous(limits = c(160, 240), breaks = c(160, 180, 200, 220, 240))+
  scale_x_continuous(limits = c(2.5,6.5))+
  scale_color_npg() #npg = nature publishing group. Also try jco, simpsons, futurama, tron. ALso see "material" 
 
#to view the whole palette and get hex codes, use the "scales" package: 

show_col(pal_npg("nrc")(12))
show_col(pal_futurama("planetexpress")(12))
show_col(pal_simpsons("springfield")(12))
show_col(pal_material("cyan")(10)) 

#8. Adding in regression lines with their corresponing equations, R-squared, and p-values
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species))+ 
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  labs(y = "Flipper Length (mm)", x = "Body Mass (kg)")+
  scale_y_continuous(limits = c(160, 240), breaks = c(160, 180, 200, 220, 240))+
  scale_x_continuous(limits = c(2.5,6.5))+
  scale_color_manual(values = c("darkorange", "darkorchid", "cyan4"))+
  stat_smooth(method = "lm", se = FALSE, size = 1.5)+
  #stat_regline_equation(label.x = 4, show.legend = FALSE)+ #this will show the equations of the regression line
  stat_cor(show.legend = FALSE)


#9. Customizing font sizes using the all-powerful theme() function. Make sure this comes AFTER theme_classic, because you are modifying components of this theme
ggplot(penguins, aes(x = mass_kg, y = flipper, color = species))+ 
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  labs(y = "Flipper Length (mm)", x = "Body Mass (kg)")+
  scale_y_continuous(limits = c(160, 240), breaks = c(160, 180, 200, 220, 240))+
  scale_x_continuous(limits = c(2.5,6.5))+
  scale_color_manual(values = c("darkorange", "darkorchid", "cyan4"))+
  stat_smooth(method = "lm", se = FALSE, size = 1.5)+
  stat_cor(show.legend = FALSE)+
  theme(axis.text.y=element_text(size=20, color= 'black'), 
        axis.text.x=element_text(size=20, color= 'black'), 
        axis.line.y=element_line(color = 'black',size=0.5),
        axis.line.x=element_line(color = 'black',size=0.5),
        axis.ticks.y=element_line(size=0.5),
        axis.ticks.x=element_line(size=0),
        axis.line = element_line(colour = "black"),
        axis.title.y=element_text(size=20),
        axis.title.x=element_text(size=20),
        legend.text = element_text(size = 18),
        legend.title = element_blank())
  
  
  
```

For the full suite of theme() customization options, check out https://ggplot2.tidyverse.org/reference/theme.html

A full list of ggpubr functions is available here: http://rpkgs.datanovia.com/ggpubr/reference/index.html

Check out all the ggsci palette options here: https://nanx.me/ggsci/articles/ggsci.html

A fantastic reference for exploring ggplot chart types is available under "chart types" here: https://www.r-graph-gallery.com/ggplot2-package.html

###4. Create a box plot
####Now we want to visualize how body size differs between the sexes for each species

Default plot:
```{r}

ggplot(penguins, aes(x = species, y = mass_kg))+
  geom_boxplot()

```

Fixed up using our new ggplot skills!
```{r}
#Option #1: open boxes, data points shown
ggplot(penguins, aes(x = species, y = mass_kg, color = sex))+
   geom_boxplot(width = 0.6, size = 1)+
  geom_point(position = position_dodge(0.6), size = 2, alpha = 0.6)+
  scale_color_manual(values = c("#EE6C00FF", "#00838EFF"),labels = c("Female", "Male"))+
  theme_classic()+
  scale_y_continuous(limits = c(2,6),breaks = c(2,3,4,5,6))+
  labs(x = "", y = "Mass (kg)")+
  theme(axis.text.y=element_text(size=20, color= 'black'), 
        axis.text.x=element_text(size=20, color= 'black'), 
        axis.line.y=element_line(color = 'black',size=0.5),
        axis.line.x=element_line(color = 'black',size=0.5),
        axis.ticks.y=element_line(size=0.5),
        axis.ticks.x=element_line(size=0),
        axis.line = element_line(colour = "black"),
        axis.title.y=element_text(size=20),
        axis.title.x=element_text(size=20),
        legend.text = element_text(size = 18),
        legend.title = element_blank())
 

#Option #2: filled boxes, no data points
ggplot(penguins, aes(x = species, y = mass_kg, fill = sex))+
   geom_boxplot(width = 0.6, size = 1)+
  theme_classic()+ 
  scale_fill_manual(values = c("#EE6C00FF", "#00838EFF"), labels = c("Female", "Male"))+
  scale_y_continuous(limits = c(2,6.5),breaks = c(2,3,4,5,6))+
  labs(y = "Mass (kg)", x = "")+
  stat_compare_means(method = "t.test", label = "p.format")+
  theme(axis.text.y=element_text(size=20, color= 'black'),
        axis.text.x=element_text(size=20, color= 'black'), 
        axis.line.y=element_line(color = 'black',size=0.5),
        axis.line.x=element_line(color = 'black',size=0.5),
        axis.ticks.y=element_line(size=0.5),
        axis.ticks.x=element_line(size=0),
        axis.line = element_line(colour = "black"),
        axis.title.y=element_text(size=20),
        axis.title.x=element_text(size=20),
        legend.text = element_text(size = 18),
        legend.title = element_blank())
        #legend.position = "top") or "none"
 

```

###5. Create a point range plot 
####For this plot, we will add points that represent mean values with error bars showing +/- standard error. We will then layer on a line plot to connet the points to show change over time.

Note: This dataset does not show strong changes over time (there are only 3 years of data) so this plot does not look very exciting. However, layering these 2 plot types can be a great way to visualize changes in mean values over time.
```{r}

penguin_means <- penguins %>% 
  group_by(species, year) %>% 
  summarise(mean_mass = mean(mass_kg), mean_flipper= mean(flipper), mean_bill_length = mean(bill_length), sd_mass = sd(mass_kg), sem = sd_mass/sqrt(n()))

ggplot(penguin_means)+
  geom_line(aes(x = year, y = mean_mass, color = species), size =1)+
  geom_pointrange(aes(x = year, y = mean_mass, ymin = mean_mass - sem, ymax = mean_mass + sem, color = species), size = 1)+
  #geom_hline(yintercept = 4.5, linetype = "dashed")+ #if you want to add horizontal or vertical lines to indicate means or asymptotes
  #geom_vline(xintercept = 2008, color = "red")+
   theme_classic()+
  labs(y = "Body Mass (kg)", x = "Year Sampled")+
  scale_x_continuous(limits = c(2007, 2009), breaks = c(2007, 2008, 2009))+
  scale_color_manual(values = c("darkorange", "darkorchid", "cyan4"))+
  theme(axis.text.y=element_text(size=20, color= 'black'), 
        axis.text.x=element_text(size=20, color= 'black'), 
        axis.line.y=element_line(color = 'black',size=0.5),
        axis.line.x=element_line(color = 'black',size=0.5),
        axis.ticks.y=element_line(size=0.5),
        axis.ticks.x=element_line(size=0),
        axis.line = element_line(colour = "black"),
        axis.title.y=element_text(size=20),
        axis.title.x=element_text(size=20),
        legend.text = element_text(size = 18),
        legend.title = element_blank())
  
  
```


BONUS: Inset plots!! patchwork::inset_element is a BRAND NEW function that allows you to create an inset plot within your main plot. 

A Messy Example:
```{r}
library(patchwork)

plot2 <- ggplot(penguins, aes(x = species, y = bill_ratio, fill = sex))+
  geom_boxplot()+
  theme_classic()+
 theme(legend.position = "none")

ggplot(penguins, aes(x = bill_length, y = bill_depth, color = species))+
  geom_point(alpha = 0.8, size = 2)+
  theme_classic()+
  scale_color_manual(values = c("darkorange", "darkorchid", "cyan4"))+
  stat_smooth(method = "lm", se = FALSE, size = 1.5)+
  inset_element(plot2, 0.6, 0.6, 1, 1, align_to = "full")

```

