---
title: "RStudio Recap"
author: "Heili Lowman"
date: "9/1/2020"
output: html_document
---

The following is an RMarkdown generated during Week 4 of SCCWRP Tidy Tuesday sessions.

Participants were instructed to bring their own datasets and work alongside the below workflow to:

1. Attach the necessary packages.
1. Import their data (in .csv form).
1. Tidy their datasets.
1. Create a final plot of their data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Demo Dataset

In the following RMarkdown, I will be using the `lynx` dataset from `data()` in base R to practice importing, examining, tidying, and visualizing a dataset.

Remember - to insert a new code chunk either click on **Insert** > **R** at the top of the screen or use one of the shortcuts: 

- Mac - cmd + opt + I
- PC - ctrl + alt + I

## Data & Packages

```{r packages, message = FALSE}

# Added "message = FALSE" to header to suppress the messages you get when loading the tidyverse printing out in your knit document.

# Load necessary packages.
library(tidyverse)

```


```{r data}

# Load in your dataset.
# data_set <- read_csv("my_data.csv")

# Call built-in dataset from base R.
lynx_data <- lynx

# Examine data. Then comment out, so it doesn't run in your script.
# View(lynx_data)

```


## Data Tidying

```{r tidy}

# Examine structure. Also commenting out so it doesn't run.
#str(lynx_data)

# My dataset is a time series, which can be a helpful format for some analyses, but I don't want it that way for this purpose.

# Make my data *not* a time series - tk_tbl() turns it into a tibble, another name for a normal dataset.
library(timetk)
lynx_df <- tk_tbl(lynx_data)
#View(lynx_df)

# Change column names.
lynx_df <- lynx_df %>% # Base dataset.
  rename(Year = index, Lynx = value) # Renames columns.

# Q & A:
# Changing column structure:
# as.numeric(dataset_name$column_name) - coerces column to be numeric, forces all non-numeric values to be "NA", may not work if you have an unrecognized character symbol
# as.character() - coerces all text (numbers and letters) to be recognized as characters
# as.factor() - also a text string, but can be categorical/ordinal
#   relevel() - relevel factors in a particular order

# To create a new column in a different format, let's pipe it using the tidyverse!
# dataset <- dataset %>%
#  mutate(new_column_name = as.numeric(old_columnn_name)) 

# It's always a good idea to keep your raw/previous data intact, and add a newly formatted column under a new name.

# Dealing with NAs:
# naniar - good package for dealing with NAs, opposite of replace_NA
# replace_NA() - replacing NAs with specified values, opposite of naniar


```


## Data Visualization

```{r plot, message = FALSE, echo = FALSE}

# I've added "echo=FALSE" to my header, because I only want my figure, not my code, to knit in the final document.

# Create a basic scatterplot of my lynx trapping data.
lynx_fig <- ggplot(lynx_df, aes(x = Year, y = Lynx)) + # basic dataset
  geom_point(size = 3, alpha = 0.75) + # creates scatterplot
  geom_smooth(method = lm, color = "aquamarine2") + # adds linear model
  labs(x = "Year",
    y = "Number of Lynx Trapped",
    title = "Lynx R Dataset",
    subtitle = "Tidy Tuesday demo, September 1",
    caption = "Data Source: Brockwell and Davis 1991") + # includes labels
  theme_classic() + # theme with only axes present
  theme(legend.position = "none") + # suppresses legend
  scale_x_continuous(breaks = seq(1820, 1940, 10)) # edits x axes tick marks
  
lynx_fig # Calls figure.  

# Q & A:
# If you'd like to add more customization, you can put in "color=" or "shape=" in the aesthetics of ggplot.
# If you're customizing multiple geometries, you should add those customizations to the parentheses of that geometry - e.g., geom_point(aes(color="aquamarine3")).
# NOTE - R accepts the British spelling of a few things, including "colour" in lieu of "color".
  
# If you'd like to comment out a bunch of code, you can use the commenting out shortcut:
# Mac: Command + Shift + C
# PC: Ctrl + Shift + C

# If you'd like to filter your dataset prior to plotting, you can again use the tidyverse to do so:
# leah_fig <- orig_data %>% # original dataset
#   filter(column_name == "A") %>% # filter for only values of "A" in a column
#   ggplot(aes()) + # plot as normal
#   geom_point()

# If you'd like to create boxplot with raw data values shown on top, you can layer your geometries like so:
# ggplot(data, aes(x = x, y = y)) +
#   geom_boxplot() + # creates boxplot
#   geom_jitter(width = 0.1) + # creates jittered points according to each category along the x axis
#   geom_hline(y = 5) + # create a horizontal line across the whole plot at y = 5
#   geom_vline(x = 2) # create a vertical line across the whole plot at x = 2

# Geometries appear in the order in which they're typed in your ggplot text, so layer logically!

```

*End of RMarkdown script.*