---
title: "Tidy Tuesday"
subtitle: "Pretty Interactive Data Viz with ggplot and plotly"
author: "Kate Culhane & Caroline Owens"
date: "06 Oct 2020"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: cerulean
    highlight: haddock
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.height = 3, fig.width = 5)
```

```{r, include = FALSE}
# The options above the first chunk are for customizing how your knitted markdown will look. You can google other themes and highlights to change it. toc creates a table of contents; the depth is the number of #'s that will be included as header levels, and `float` lets the TOC follow you as you scroll down.
```
<br>

*Heads up about the way I code:*

- I like displaying plots in the R Studio console instead of inline (you can switch this setting for each document using the gear)
- I'll try to use double quotations (") instead of single ('), but in case I forget...they're the same!
- Keyboard shortcuts: https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts
  - **Run code**
    - ctrl/cmd + enter
  - **%>%**
    - ctrl/cmd + shift + m
  - **Insert code chunk**
    - ctrl/cmd + alt + i
  - **Comment out lines**
    - ctrl/cmd + shift + c
    
```{r}

# Load packages
library(tidyverse) # readr (read_csv and other data loading functions), dplyr (select, filter, mutate, and other data wrangling functions), magrittr (the piping function), ggplot2 (everything dataviz)
library(plotly) # making interactive graphs (ggplotly())
library(htmlwidgets) # Exporting graphs as HTML (saveWidget())

# Load data
penguins <- read_csv("penguins.csv")

# You can also do the following to load the data via CRAN:
# install.packages("palmerpenguins")
# library(palmerpenguins)
# penguins <- penguins

```

<br>

## 1: IncrediBILL Palmer Penguins

**Same data set as last week: Palmer Penguin dataset from LTER data.**

### Data exploration

```{r}

# I like to run all these in the console or comment them out after running 

#View(penguins) # Can also click on the object in the global environment
str(penguins) #shows the structure of the dataset
tail(penguins, 5) #shows the last 5 rows of the dataset

```

<br>

### Static scatterplot

**REVIEW:** Create a scatterplot using `ggplot`.

- x-axis: `body_mass_g`
- y-axis: `bill_length_mm`
- point color: `species`

*Hint:* Use `ggplot`, `geom_point`, and `aes(color = "")`

Tips:

- `ggplot2` is the package, `ggplot` is the function
- Things ggplot HAS to know: 1) dataset, 2) aesthetics, 3) geom
- Dataset and aes can be specified either in `ggplot()` or any `geom_*`
- Aes is for visual changes that depend on a value in the data (as opposed to visual changes that apply to all values)

```{r}

ggplot(data = penguins,
       aes(x =  body_mass_g, y = bill_length_mm,
           color = species, shape = sex)) +
  geom_point(na.rm = TRUE) # suppresses warning code about removed NAs

```

<br>

#### Beautification

Tips:

- R Studio cheatsheets: https://rstudio.com/resources/cheatsheets/
- Order doesn't matter... as long as you name your vectors
  - When adjusting scales, R will apply your color/shape values in the same order as levels in your data
  - If you name the elements of your scale vector after your data values, R will match values instead of using order - this is more foolproof!
- Choosing color schemes
  - Colors in R: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
  - Hex codes: https://coolors.co/ (you can also upload a photo to match colors from)
  - Think about data type: continuous, discrete, ordinal (ordered), diverging
- Choosing shapes
  - In R: http://www.sthda.com/english/wiki/ggplot2-point-shapes
  - Unicode: https://jrgraphix.net/r/Unicode/2600-26FF
- Adjusting themes
  - Changing elements of a theme: https://ggplot2.tidyverse.org/reference/theme.html

```{r}

# Make the plot pretty!

# Save a vector of colors for use in future graphs
col_scale <- c("#A3E7FC", "#32908F", "#26C485") # c() combines values into a vector
names(col_scale) <- c("Adelie", "Gentoo", "Chinstrap") # naming each value to correspond to the species

# Save a vector of names for use in the legend
lab_scale <- c("P. adeliae", "P. papua", "P. antarcticus") #scientific names :)
names(lab_scale) <- c("Adelie", "Gentoo", "Chinstrap") #ALWAYS name your vectors and double check the order

# New pretty graph
p <- penguins %>% 
  na.omit() %>% # removes ALL lines with NAs from dataset
  ggplot(aes(x = body_mass_g/1000, # can divide by 1000 to make units more manageable
             y = bill_length_mm,
             color = species, # point color depends on species
             shape = sex)) + # point shape depends on sex
  geom_point(size = 1.5, # adjust point size
             stroke = 1.1) + # adjust outline line width
  # scale_color_manual(values = "red", "yellow", "green")) + # DOESN'T WORK
  # scale_color_manual(values = c("lightskyblue", "slateblue4", "darkslategray4")) + 
  scale_color_manual(values = col_scale,
                     labels = lab_scale, # order matters (unless you use a named vector!)
                     name = "Species") +
  # scale_shape_manual(values = c("\u2600", "\u2601")) + # using unicode values
  # scale_shape_manual(values = c("\u2640", "\u2642")) + #another unicode option
  scale_shape_manual(values = c(19, 21), #base R shapes
                     name = "Sex") + #legend title
  theme_minimal() + # one of the default themes
  theme(text = element_text(color = "grey80"), # ALL text color
        title = element_text(face = "bold"), # all TITLES to bold
        plot.title = element_text(color = "lightblue", # PLOT TITLE color
                                  size = 16, # PLOT TITLE text size
                                  hjust = 0.5, vjust = 5), # adjusting title positioning
        plot.margin = margin(25,15,10,10), # remember TRBL (trouble)
        axis.text = element_text(color = "grey70"), # AXIS LABELS color
        legend.text = element_text(face = "italic"), # LEGEND LABELS color
        legend.background = element_rect(fill = "grey25", # LEGEND BACKGROUND color
                                         color = NA), # no legend outline
        plot.background = element_rect(fill = "grey20"), # entire background color
        panel.grid = element_line(color = "grey60")) + # GRID LINE color
  labs(x = "Body mass (kg)", # Make sure units match!!! Especially after converting
       y = "Bill length (mm)",
       title = "IncrediBILL Palmer Penguins")

p #view the named plot

```

<br>

#### Saving a theme

Like your theme and want to use it for another graph without copy-pasting? Save it!

- `theme_set` replaces the default
- `theme_update` adds new elements to the default theme (analagous to `+`)
- `theme_replace` *replaces* elements of the default theme

```{r}

# What's the default theme in your R session?
ggplot(cars, aes(x = speed, y = dist)) + geom_point()

# Set the default as one of the included themes
theme_set(theme_classic())
#Check it out
ggplot(cars, aes(x = speed, y = dist)) + geom_point()

# Make your own theme!
theme_awful <- theme_dark() # save a copy of a default theme with a new name
theme_set(theme_awful) # set your copy as the default
theme_update(panel.grid.minor = element_line(colour = "red", size = 3), #update your copy!
             plot.background = element_rect(fill = "yellow"),
             title = element_text(size = 60, face = "italic", color = "tan"),
             axis.text = element_text(color = "purple", face = "bold", angle = 160))

#Check your new default theme out
ggplot(cars, aes(x = speed, y = dist)) + geom_point()

```

<br>

#### Saving the plot

- Better to use code than the "export" button
  - Can control output size and keep values consistent
  - Whatever you do...don't screenshot! Gives you terrible resolution.

```{r}

# Check filepath
getwd() # usually run in the console - check where on your computer the plot will be saved

ggsave("penguin_plot.png", # filepath will be added to WD path (prob just filename) - INCLUDE FILE TYPE EXTENSION.
       plot = p, # name of object to save - default is last_plot()
       height = 4, width = 5, units = "in") # output size (can use inches, cm, pixels)

```

<br>

### Interactive scatterplot

Plotly

- Plotly is a platform that interacts with R, Python, and Julia
- Plotly focuses on creating interactive data visualizations and web applications (check out "Dash")
- We'll be using ggplotly, which easily turns ggplots into interactive graphs: https://plotly.com/ggplot2/

```{r}

ggplotly(p) # notice that the plotly graph displays in the "Viewer" pane!

```

<br>

## 2: Flippin' Palmer Penguins

### Static scatterplot

**REVIEW:** Create a scatterplot using `ggplot`.

- x-axis: `flipper_length_mm`
- y-axis: `body_mass_g`
- point color: `species`, same color scheme as above (use the saved vector!)

```{r}

theme_set(theme_classic()) # switching defaults since I'm not a huge fan of theme_awful...

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g,
                     color = species)) +
  geom_point(na.rm=TRUE) + # suppress warning about NAs being removed
  scale_color_manual(values = col_scale) # same color vector as previous graph!

```

<br>

#### Quick beautification

Tips:

- Always make a static graph first - but don't spend time optimizing all the theme elements!
  - some formatting elements do not translate
  - legends, element sizes, and themes get funky with ggplotly

The following aesthetic parameters will stay nice when you run your graph through Plotly:
```{r}

ggplot(penguins, aes(x = flipper_length_mm/10, y = body_mass_g/1000, # adjust units
                           fill = species)) + # changes INSIDE color of the point
  geom_point(na.rm = TRUE,
             shape = 21, # this particular shape allows for a color AND a fill
             color = "grey30") + # changes OUTSIDE color of the point
  scale_fill_manual(values = col_scale, # use scale_fill_* because we used "fill" in the aes
                    name = "Species") +
  labs(x = "Flipper length (cm)", # match units!
       y = "Body mass (kg)", # match units!
       title = "Flippin' Palmer Penguins")

```

<br>

### Interactive scatterplot

```{r}

# Updated static plot
p2 <- ggplot(penguins, aes(x = flipper_length_mm/10, y = body_mass_g/1000,
                           fill = species,
                           cooltip = sex)) + # text to use for tooltip - can have any name
  geom_point(na.rm = TRUE,
             size = 2, shape = 21, color = "grey30") +
  scale_fill_manual(values = col_scale,
                    name = "Species") +
  theme_classic() +
  labs(x = "Flipper length (cm)",
       y = "Body mass (kg)",
       title = "Flippin' Palmer Penguins")

# Interactive plot
ggplotly(p2,
         tooltip = "cooltip") # changes tooltip text. Remember - you can add any data column to the tooltip as long as you name it in aes()

```

<br>

### Animated scatterplot

This is ideal for timeseries data - especially measurements that are paired between years.
```{r}

# Updated static plot
p3 <- ggplot(penguins, aes(x = flipper_length_mm/10, y = body_mass_g/1000,
                           fill = species,
                           cooltip = sex,
                           frame = year)) + # adds animation, with one frame per year
  geom_point(na.rm = TRUE,
             size = 2, shape = 21, color = "grey30") +
  scale_fill_manual(values = col_scale,
                    name = "Species") +
  theme_classic() +
  labs(x = "Flipper length (cm)",
       y = "Body mass (kg)",
       title = "Flippin' Palmer Penguins")

# Interactive plot
p3_move <- ggplotly(p3,
                    tooltip = "cooltip")

p3_move

```

<br>

#### Saving the plot

*Save as HTML*

- Can save in the HTML output of R markdown
- Can save as a standalone HTML file using `htmltools`

```{r}

# Save as standalone HTML
saveWidget(p3_move, "penguin_animation.html") #this function is from the htmlwidget package

```

*Export to plotly account*

- Create a free plotly account: chart-studio.plotly.com
  - Find your username and API key
- Embed the plot in a webpage: https://help.plotly.com/embed-graphs-in-websites/

```{r, message = FALSE, output = FALSE}

# Save plotly username and API key
Sys.setenv("plotly_username" = "katekathrynkat")
Sys.setenv("plotly_api_key" = "WmLiHIWtsIoNpfMz96a6")

# Explort plot to plotly
api_create(x = p3_move, # name of plot
           fileopt = "overwrite", sharing = "public")

```

See Kate's embeddable graph: https://chart-studio.plotly.com/~katekathrynkat/39/#/

-Scroll to bottom right
-There are 4 icons: Facebook, Twitter, link, <>
-The <> icon will bring up the code you can give to a webmaster or embed on an html page

