
---
title: "leaflet tutorial"
author: "Maggie Klope"
date: "1/12/2021"
output: html_document
---

## Leaflet: Easily create interactive maps

- Uses the [Leaflet](https://leafletjs.com/index.html) open-source JavaScript library
- Quick way to make interactive maps
- Can be embedded into R Markdown and Shiny apps
- Compatible with a lot of data types (data frames with lat/long columns, sp and sf package spatial objects)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install packages if you have not already
# install.packages("tidyverse")
# install.packages("leaflet")
# install.packages("leaflet.minicharts")
# install.packages("sf")

library(tidyverse)
library(leaflet)
library(leaflet.minicharts)
library(sf)

```

### Basic Map with one point
```{r}

# Using the lat/long for the Methuselah Tree (oldest non-clonal tree in the world!)
methuselah <- leaflet() %>% # using the '%>%' pipe from tidyverse
  addTiles() %>% # loads default baselayer map
  addMarkers(lat = 37.379444, 
             lng = -118.161667, 
             label = "Oldest Tree!",
             popup = "Methuselah Tree") # popup will add a small window that appear over the point

methuselah

# adding more features
methuselah %>% 
  addProviderTiles(providers$Esri.WorldStreetMap) %>% # changes basemap
  addMiniMap() %>% # add in a second zoomed out map %>% 
  addScaleBar()

# lots of different privider tiles: https://leaflet-extras.github.io/leaflet-providers/preview/
methuselah %>% addProviderTiles(providers$Esri.WorldImagery)

methuselah %>% addProviderTiles(providers$Stamen.Watercolor)
  
```

### Using shapefile

- Audubon Society occurrence surveys
- Includes a count of sensitive and threatened species
- Assigns a Score value for the taxa that exceed the population threshold of 1% of global or 10% of state population

```{r}

# use sf to read the shapefile
birds <- st_read("ds78.shp") %>% 
  filter(LONGITUDE != 0.0000) %>%  # removing a few data points that were missing lat/long
  filter(REGION == "Southwest")

# creating a point map
bird_map <- leaflet(data = birds) %>% 
  addTiles() %>% 
  addMarkers(lat = birds$LATITUDE, lng = birds$LONGITUDE)

```

### Creating a cluster map
```{r}
leaflet(data = birds) %>%
  addTiles() %>%
  addMarkers(lat = birds$LATITUDE, 
             lng = birds$LONGITUDE,
             clusterOptions = markerClusterOptions(),
             label = birds$NAME)

```

### Creating custom icons
```{r}

# making a dataset for locations where brown pelicans (BRPN) were detected
pelicans <- birds %>% 
  filter(BRPN == "Y") # filtering where the BRPN column has a "Y" for yes

# using the makeIcon() function to load a .png file
pelican_icon <- makeIcon(iconUrl = "pelican.png", iconHeight = 20, iconWidth = 20)

leaflet(data = pelicans) %>% 
  addTiles() %>% 
  addMarkers(lat = pelicans$LATITUDE, 
             lng = pelicans$LONGITUDE, 
             icon = pelican_icon) # setting the icon argument to the png that we loaded

```

### Using the leaflet.minicharts package, you can add figures over your map	
```{r}	
bird_matrix <- as.tibble(birds)

# adding barcharts	
leaflet(data = bird_matrix) %>% 	
  addTiles() %>% 	
  addMinicharts(	
    lng = bird_matrix$LONGITUDE,	
    lat = bird_matrix$LATITUDE,	
    chartdata = bird_matrix[, c("SCORE", "THREAT")],	
    type = "bar") # just using the Score and Threat columns	

# adding piecharts	
leaflet(data = bird_matrix) %>% 	
  addTiles() %>% 	
  addMinicharts(	
    lng = bird_matrix$LONGITUDE,	
    lat = bird_matrix$LATITUDE,	
    chartdata = bird_matrix[, c("SCORE", "THREAT")],	
    type = "pie")	

# adding label	
leaflet(data = bird_matrix) %>% 	
  addTiles() %>% 	
  addMinicharts(	
    lng = bird_matrix$LONGITUDE,	
    lat = bird_matrix$LATITUDE,	
    chartdata = bird_matrix$SCORE,	
    showLabels = TRUE)	
```	

### Maps with Polylines	

```{r}	
# loading example dataset from the leaflet package	
storm_data <- atlStorms2005	

# Create a continuous palette function	
palette <- colorNumeric(	
  palette = "Reds",	
  domain = storm_data$MaxWind)	

# Creating a map
storm <- leaflet(storm_data) %>% 	
  addTiles() %>% 	
  # addPolylines(color = "red") %>% 	
  addPolylines(color = ~palette(MaxWind), opacity = .5) %>% 	
  addLegend(pal = palette, values = ~MaxWind, 	
            title = "Maximum Wind", 	
            opacity = .5)	

storm	

# some extra features	
storm %>% 	
  addGraticule() %>% 	
  addTerminator()	
```	

### Compatible with Shiny	

- See the leaflet_shiny_app.R file